---
title: "Homicide rate, Suicide rate and GDP over 50 years"
subtitle: "Introduction to R for Data Science and Computational Social Sciences"
author: "Ursulina Kölbener, Cand. Lucerne Master in Computational Social Sciences"
date: "2022-11-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup
```{r packages, echo = FALSE}
# install and load packages
# install.packages("tidyverse")
library(tidyverse)
```


## Dataset

The dataset contains homicide rate, suicide rate and gross domestic product (GDP) from 1970 to 2020 by country. The data source is [THE WORLD BANK](https://data.worldbank.org/).

```{r import data}
# import dataset
shg_original <- rio::import("data/suicide_homicide_gdp.csv")
```

There are 13'566 observations with 10 variables. The country name is represeneted in three different ways. Once completely written out (`country`), once abbreviated according to the iso standard with 3 characters (`iso3c`) and once abbreviated according to the iso standard with 2 characters (`iso2c`).

```{r observations and variables}
# number of observations
nrow(shg_original)

# numer of variables
ncol(shg_original)

```

The dataset shows observations from 1970 to 2020 (`year`) as a integer data type. Further there are four numerical values: `Intentional homicides (per 100,000 people)`, `Suicide mortality rate (per 100,000 population)`, Gross Domestic Product `GDP (current US$)` and `GDP per capita, PPP (current international $)`. Intentional homicides and suicide mortality rate are discrete values. Gross domestic product (GDP) is a continuous variable, a monetary measure of the market value of all the final goods and services produced and sold (not resold) in a specific time period by countries. Gross domestic product (GDP) per capita is a financial metric that breaks down a country's economic output per person and is calculated by dividing the GDP of a nation by its population. The values of the GDP per capita reach from 285.4 to 153563.9.

Finally, the region (`adminregion`) and income level (`incomeLevel`) are specified as characters. Region and income level are categorical variables. Income level is somewhat ordinal with the following values: Aggregates, High income, Low income, Lower middle income	and Upper middle income. A few variables are not classified, others are aggregates. I will remove them later.

```{r data type and summary}
# check data type of every variable
str(shg_original)

# summary statistics
summary(shg_original)

```

### Research Questions

Do intentional homicides correlate with gross domestic product (GDP) in genereal? What are the differences in regions over the years?

Does a high homicide rate favor a high suicide mortality rate?


### Data Transformation

Transforming the data for easier handling. Also, the dataset contains lot's of variables with NA. Removing those NA's leaves 2352 observations.
```{r}
# creating a new dataset from the original data
shg <- shg_original

# renaming the variables to easier names
shg <- rename(shg, IntHomicides = `Intentional homicides (per 100,000 people)`,
              SuicideMort = `Suicide mortality rate (per 100,000 population)`,
              GDP = `GDP (current US$)`,
              GDPcap = `GDP per capita, PPP (current international $)`,
              region = `adminregion`)

# removing all observations with NA's
shg <- shg %>% 
  filter(incomeLevel != "Aggregates" & incomeLevel != "Not classified" ) %>% 
  filter(!is.na(IntHomicides)) %>% 
  filter(!is.na(SuicideMort))

```


## Summary Statistics

For the *income level*: There are 897 observations with high income, 757 with upper middle income, 560 with lower middle income and 132 observations with low income.

```{r}
# calculate frequency of income level
table(shg$incomeLevel)
```

GDP per capita: The minimum value is 532.5 and the maximum 141634.7. The mean has a value of 19491.1, the medain is lower at 12506.5.
```{r}
# summary of GDP per capita
summary(shg$GDPcap)
```


```{r}
# average intention homicides, suicidate mortality and GDP per capita
# grouped by income level
shg %>% 
  group_by(incomeLevel) %>% 
  summarise(avg_IntHomicides = mean(IntHomicides, na.rm = TRUE),
            avg_SuicideMort = mean(SuicideMort, na.rm = TRUE),
            avg_GDPcap = mean(GDPcap, na.rm = TRUE))
```

## Data Transformation
```{r}
# create a new subset
shg_trnsf <- shg %>% 
  select(-iso3c) %>% # remove column iso3c
  filter(iso2c != "") # remove rows (total per region)

```


As the regions in the dataset are incomplete and do not match the ISO standard, I imported a proper list (`region`) with countries and their regions. I merged the two tables together. I used `merge`as I could not figure it out with `inner_join`.
```{r}
# import a second dataset
region <- rio::import("data/region.csv")

# rename columns for easier merging
region <- rename(region, iso2c = `alpha-2`,
                 regionISO = `region`)

# combining another dataset
shg_exp <- merge(x = shg_trnsf, y = region, by = "iso2c")

```


## Data Vizualization
### Global Trend

```{r}
# calculate the average of three continuous variables
shg_trend <- shg %>% 
  group_by(year) %>% 
  summarise(avg_IntHomicides = mean(IntHomicides, na.rm = TRUE),
            avg_SuicideMort = mean(SuicideMort, na.rm = TRUE),
            avg_GDPcap = mean(GDPcap, na.rm = TRUE))

# show summarized data
shg_trend # lots of NaN in there, because there is missing data for every country in a year

# remove rows with missing data
shg_trend_woNaN <- na.omit(shg_trend) # only 19 years of complete data left

```

```{r}
# plot Global Trend in Global Trend in GDP per Capita" 
ggplot(data = shg_trend[!is.na(shg_trend$avg_GDPcap),]) +
  geom_line(mapping = aes(x = year, y = avg_GDPcap)) +
  theme_light() +
  labs(x = "year", y = "GDP per capita (current international $)", 
       title = "Global Trend in GDP per Capita",
       subtitle = "",
       caption = "Data from the world bank.")
```

```{r}
# plot Global Trend in Intentional Homicides 
ggplot(data = shg_trend[!is.na(shg_trend$avg_IntHomicides),]) +
  geom_line(mapping = aes(x = year, y = avg_IntHomicides)) +
  theme_light() +
  labs(x = "year", y = "Intentional homicides (per 100,000 people)", 
       title = "Global Trend in Intentional Homicides",
       subtitle = "",
       caption = "Data from the world bank.")
  
```

```{r}
# plot Global Trend in Global Trend in Suicide Mortality Rate 
ggplot(data = shg_trend[!is.na(shg_trend$avg_SuicideMort),]) +
  geom_line(mapping = aes(x = year, y = avg_SuicideMort)) +
  theme_light() +
  labs(x = "year", y = "Suicide mortality rate (per 100,000 population)", 
       title = "Global Trend in Suicide Mortality Rate",
       subtitle = "",
       caption = "Data from the world bank.")
  
```

```{r}
# plot Global Trend 
ggplot(data = shg_trend_woNaN, mapping = aes(x = year)) +
  geom_line(mapping = aes(y = avg_IntHomicides), colour = "#33638DFF", size = 1.2) +
  geom_line(mapping = aes(y = avg_SuicideMort), colour = "#29AF7FFF", size = 1.2) +
  theme_light() +
  labs(x = "year", y = "cases per 100,000 population", 
       title = "Global Trend in Suicide Mortality Rate",
       subtitle = "",
       caption = "Data from the world bank.")
  
```

```{r, echo = FALSE}
# remove NA's
shg_woNA <- na.omit(shg) %>% 
  filter(region != "")

ggplot(data = shg_woNA) +
  geom_point(mapping = aes(x = log(GDPcap), y = IntHomicides))+
  geom_smooth(mapping = aes(x = log(GDPcap), y = IntHomicides), method = "lm")+
  facet_wrap (~region, scales = "free") +
  theme_light()

```

```{r}
ggplot(data = shg_woNA) +
  geom_point(mapping = aes(x = log(GDPcap), y = SuicideMort, colour = region))+
  theme_light()

```
## Statistical Modelling
```{r}
# study Suicide Mortality Rate
fit1 <- lm(formula = SuicideMort ~ IntHomicides + GDPcap, 
           data = shg_exp)

# show results
fit1
```


